---
title: "Cyclistic Data Analysis Project - Full Report"
author: "Hazem El-Mancy"
date: "`r Sys.Date()`"
output: html_document
---
> ## Overview

This report is a part of the [“Google Data Analytics Professional Certificate’s”](https://www.coursera.org/professional-certificates/google-data-analytics) capstone 
project offered by [Coursera](https://www.coursera.org/).
In this report, the chosen case study is the first problem in track#1, also known as 
‘Cyclistic,’ an imaginary bike-sharing company in Chicago. The company offers both 
traditional and assistive bikes to both casual riders and riders who are annual members.  

> ## Scenario

“You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations”.

> ## Business Task

The questions that the stakeholders want answer in this scenario are: 

* How do annual members and casual riders use Cyclistic bikes differently?   

* Why would casual riders buy Cyclistic annual memberships?   

* How can Cyclistic use digital media to influence casual riders to become members?   

From these questions, the priority-1 task in this project is to identify the differences of bike 
usage between annual members and casual riders. Then, use those differences to figure out 
how to keep the existing and attract new annual members to maximize Cyclistic’s profit.

> ## Data Preparation

Data used is the last 12 months data of ride trips made by casual and member riders. The links to data are below:

* [March-2021](https://divvy-tripdata.s3.amazonaws.com/202103-divvy-tripdata.zip)

* [April-2021](https://divvy-tripdata.s3.amazonaws.com/202104-divvy-tripdata.zip)

* [May-2021](https://divvy-tripdata.s3.amazonaws.com/202105-divvy-tripdata.zip)

* [June-2021](https://divvy-tripdata.s3.amazonaws.com/202106-divvy-tripdata.zip)

* [July-2021](https://divvy-tripdata.s3.amazonaws.com/202107-divvy-tripdata.zip)

* [August-2021](https://divvy-tripdata.s3.amazonaws.com/202108-divvy-tripdata.zip)

* [September-2021](https://divvy-tripdata.s3.amazonaws.com/202109-divvy-tripdata.zip)

* [October-2021](https://divvy-tripdata.s3.amazonaws.com/202110-divvy-tripdata.zip)

* [November-2021](https://divvy-tripdata.s3.amazonaws.com/202111-divvy-tripdata.zip)

* [December-2021](https://divvy-tripdata.s3.amazonaws.com/202112-divvy-tripdata.zip)

* [January-2022](https://divvy-tripdata.s3.amazonaws.com/202201-divvy-tripdata.zip)

* [February-2022](https://divvy-tripdata.s3.amazonaws.com/202202-divvy-tripdata.zip)

Now, we can open RStudio and create new script and start data analysis journey.

### Load libraries

```{r Load Libraries}
library(tidyverse)
library(readr)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(lubridate)
library(janitor)
library(skimr)
```

### Import Data

```{r Import Data}
Mar_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202103-divvy-tripdata.csv"
  )
Apr_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202104-divvy-tripdata.csv"
  )
May_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202105-divvy-tripdata.csv"
  )
Jun_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202106-divvy-tripdata.csv"
  )
Jul_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202107-divvy-tripdata.csv"
  )
Aug_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202108-divvy-tripdata.csv"
  )
Sep_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202109-divvy-tripdata.csv"
  )
Oct_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202110-divvy-tripdata.csv"
  )
Nov_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202111-divvy-tripdata.csv"
  )
Dec_2021 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202112-divvy-tripdata.csv"
  )
Jan_2022 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202201-divvy-tripdata.csv"
  )
Feb_2022 <-
  read.csv(
    "C:/Users/hazem/Desktop/Case-Studies/Bike-Share/data/202202-divvy-tripdata.csv"
  )
```

> ## Data Processing

First, let's have a look on what we have.

```{r Data Exploration}
glimpse(Mar_2021)
```

Since we are going to bind all months data in one data frame, it is important that all data frames columns have the same name and datatype. So, we can use the code below to compare them first.

```{r df comparison}
compare_df_cols(
  Mar_2021,
  Apr_2021,
  May_2021,
  Jun_2021,
  Jul_2021,
  Aug_2021,
  Sep_2021,
  Oct_2021,
  Nov_2021,
  Dec_2021,
  Jan_2022,
  Feb_2022,
  return = "mismatch"
)
```

We can follow the following steps to make columns specs identical and check again if there is mismatch before binding.

```{r clean names, include=FALSE}
clean_names(Mar_2021)
clean_names(Apr_2021)
clean_names(May_2021)
clean_names(Jun_2021)
clean_names(Jul_2021)
clean_names(Aug_2021)
clean_names(Sep_2021)
clean_names(Oct_2021)
clean_names(Nov_2021)
clean_names(Dec_2021)
clean_names(Jan_2022)
clean_names(Feb_2022)
```

All datatypes are perfect except start and end datetime. Let's convert them to date.

```{r date time format}
Mar_2021 <-
  mutate(Mar_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Apr_2021 <-
  mutate(Apr_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

May_2021 <-
  mutate(May_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Jun_2021 <-
  mutate(Jun_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Jul_2021 <-
  mutate(Jul_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Aug_2021 <-
  mutate(Aug_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Sep_2021 <-
  mutate(Sep_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Oct_2021 <-
  mutate(Oct_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Nov_2021 <-
  mutate(Nov_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Dec_2021 <-
  mutate(Dec_2021,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Jan_2022 <-
  mutate(Jan_2022,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))

Feb_2022 <-
  mutate(Feb_2022,
         started_at = as_datetime(started_at),
         ended_at = as_datetime(ended_at))
```

Check if there any duplicates to be removed

```{r duplicates check}
Mar_2021[duplicated(Mar_2021)]

Apr_2021[duplicated(Apr_2021)]

May_2021[duplicated(May_2021)]

Jun_2021[duplicated(Jun_2021)]

Jul_2021[duplicated(Jul_2021)]

Aug_2021[duplicated(Aug_2021)]

Sep_2021[duplicated(Sep_2021)]

Oct_2021[duplicated(Oct_2021)]

Nov_2021[duplicated(Nov_2021)]

Dec_2021[duplicated(Dec_2021)]

Jan_2022[duplicated(Jan_2022)]

Feb_2022[duplicated(Feb_2022)]
```

There are no duplicates. Now compare data frames again and bind them into one data frame.

```{r df compare}
compare_df_cols(
  Mar_2021,
  Apr_2021,
  May_2021,
  Jun_2021,
  Jul_2021,
  Aug_2021,
  Sep_2021,
  Oct_2021,
  Nov_2021,
  Dec_2021,
  Jan_2022,
  Feb_2022,
  return = "mismatch"
)
```

Now, we can bind all data frames together into one data frame.

```{r bind df}
last_12_months_trips <- bind_rows(
  Mar_2021,
  Apr_2021,
  May_2021,
  Jun_2021,
  Jul_2021,
  Aug_2021,
  Sep_2021,
  Oct_2021,
  Nov_2021,
  Dec_2021,
  Jan_2022,
  Feb_2022
) 
head(last_12_months_trips)
```

Since rideable_type and member_casual are categorical, we can transform them to factors.

```{r transform to categorical}
last_12_months_trips <- last_12_months_trips %>%
  mutate(rideable_type = as.factor(rideable_type)) %>%
  mutate(member_casual = as.factor(member_casual))
```

We have to create new columns that will help us to answer questions. These columns will give us information about each trip month, day of the week, hour of day, and time duration.

```{r add new columns}
last_12_months_trips$day_of_week <-
  format(as.Date(last_12_months_trips$started_at), "%A")

last_12_months_trips$month <-
  format(as.Date(last_12_months_trips$started_at), "%b")

last_12_months_trips$hour_of_day <-
  format(last_12_months_trips$started_at, format = "%H")

last_12_months_trips$ride_duration <-
  difftime(last_12_months_trips$ended_at,
           last_12_months_trips$started_at)
```

Convert ride duration to numeric to be shown as seconds.

```{r ride duration to seconds}
last_12_months_trips <-
  mutate(last_12_months_trips, ride_duration = as.numeric(ride_duration))
```

Check if there are ride duration < 0

```{r count illogical ride duration}
count(last_12_months_trips, ride_duration < 0)
```

Exclude illogical ride duration

```{r}
last_12_months_trips <-
  filter(last_12_months_trips, ride_duration >= 0)
```

Convert ride duration from seconds to minutes

```{r}
last_12_months_trips <-
  mutate(last_12_months_trips, ride_duration = (ride_duration / 60))
```


Now, we can use `last_12_months_trips` data frame and start analysis and answer questions.

> ## Data Analysis & Visualization

First, we can use summary function and try to get thoughts about data

```{r summary}
summary(last_12_months_trips)
```

Number of trips per month:

```{r No. of trips per month}
trips_per_month <- last_12_months_trips %>%
  group_by(month) %>%
  summarise(number_of_rides = n()) %>%
  arrange(desc(number_of_rides))
trips_per_month
```

### How do annual members and casual riders use Cyclistic bikes differently?

* According to month of the year:

```{r rides per month}
ggplot(data = last_12_months_trips,
       mapping = aes(x = month, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(title = "No. Of Rides vs Months",
       x = "Month",
       y = "Number Of Rides",
       fill = "Subscription Type") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(
    limits = c(
      "Jan",
      "Feb",
      "Mar",
      "Apr",
      "May",
      "Jun",
      "Jul",
      "Aug",
      "Sep",
      "Oct",
      "Nov",
      "Dec"
    )
  )
```

It seems that number of trips is high goes up during summer. During summer months (June, July, and August), casual rides exceeds member rides.

* According to day of the weeks:

```{r rides per days of the week}
ggplot(data = last_12_months_trips,
       mapping = aes(x = day_of_week, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(title = "No. Of Rides vs Day Of The Week",
       x = "Day Of The Week",
       y = "Number Of Rides",
       fill = "Subscription Type") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(
    limits = c(
      "Monday",
      "Tuesday",
      "Wednesday",
      "Thursday",
      "Friday",
      "Saturday",
      "Sunday"
    )
  )
```

Member rides almost the same all days of the week. Unlikely, casual rides rise up and even more than member rides during weekends. This confirms that Cyclistic users are more likely to ride for leisure.

* According hour of the day:

```{r rides per hour of the day}
ggplot(data = last_12_months_trips,
       mapping = aes(x = hour_of_day, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(title = "No. Of Rides vs Hour Of Day",
       x = "Hour Of Day",
       y = "Number Of Rides",
       fill = "Subscription Type") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(
    limits = c(
      "04",
      "05",
      "06",
      "07",
      "08",
      "09",
      "10",
      "11",
      "12",
      "13",
      "14",
      "15",
      "16",
      "17",
      "18",
      "19",
      "20",
      "21",
      "22",
      "23"
    )
  )
```

Member rides are high during rush hours which confirms the fact that about 30% of users use bikes to commute to work each day.

* According to bike type:

```{r rides per bike type}
ggplot(data = last_12_months_trips,
       mapping = aes(x = rideable_type, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(title = "Bike Type vs Subscription Type",
       x = "Bike Type",
       y = "Number Of Rides",
       fill = "Subscription Type") +
  scale_x_discrete(limits = c("docked_bike",
                              "classic_bike",
                              "electric_bike"))
```

Classic bike is the most popular bike for both casual and member riders. Members do not use docked bike.


### Casual Rides vs Member Rides:

```{r casual vs member}
summary(last_12_months_trips$member_casual)
```

Percentage in piechart:

```{r casual vs member piechart}
piechart <- data.frame(
  subscription_type = c("casual", "member"),
  value = summary(last_12_months_trips$member_casual)
)

ggplot(data = piechart,
       mapping = aes(x = "", y = value, fill = subscription_type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Casual Rides vs Member Rides")+
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    legend.position = "none"
  ) +
  annotate(
    "text",
    label = "Member",
    x = 1,
    y = 900000,
    size = 6
  ) +
  annotate(
    "text",
    label = "3,127,207",
    x = 1,
    y = 1250000,
    size = 6
  ) +
  annotate(
    "text",
    label = "55.17%",
    x = 1,
    y = 1600000,
    size = 6
  ) +
  annotate(
    "text",
    label = "Casual",
    x = 1,
    y = 4750000,
    size = 6
  ) +
  annotate(
    "text",
    label = "2,540,634",
    x = 1,
    y = 4400000,
    size = 6
  ) +
  annotate(
    "text",
    label = "44.83%",
    x = 1,
    y = 4050000,
    size = 6
  )
```

Member rides are more than casual rides.

### Top high number of rides start stations:

```{r top high start stations}
start_stations <- last_12_months_trips %>%
  filter(!start_station_name == "") %>%
  group_by(start_station_name) %>%
  summarise(number_of_rides = n()) %>%
  arrange(desc(number_of_rides))

head(start_stations, n = 10)
```

These top stations are highly recommended to start the marketting plan.


> ## Recommendations:

In last sections we answered the first question of the business task and explained how annual members and casual riders use Cyclistic bikes differently. In this sectition we are going to use insights to give recommendations that could answer the last two questions. Top recommendations are listed below.


* Since casual riders tend to rent Cyclistic bikes during summer months, we can make a marketing plan with offers and discounts to target them before and during these months.


* We don't have enough information to know if the increase in number of rides in summer is because of tourists that visit Chicago in these months. If we have home city addresses of riders, we can use them to decide if we can new branches in these cities. As they likes our bikes, by these new branches, we can convince them to be annual members.


* Marketing plan should start with stations of top rides.
```{r recommendations-stations, echo=FALSE}
head(start_stations, n = 10)
```


* Since the aim is to convert casual riders to annual members, we can set up a new price list with flexible prices plan to convince casual riders to be annual members.